Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-09-07 concurrency/CPP 2024/paper.tex
fix breakage
edits throughout

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{Node}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Node\_ref }}}
\newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
\newcommand{\infp}{\ensuremath{\mathsf{Node\_ref }}}

@@

Our specific contributions are:
\begin{itemize}
	\item We reimplement the template approach without flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/subtree''.
	\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
	\item We highlight several questions about the boundaries of the template/data structure division, and clarify an element of the template approach that has not been formally addressed.
\end{itemize}

\subsection*{Related Work}
Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}. The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach. Some of the more complex data structures verified include the \lstinline{ConcurrentSkipListMap} from \lstinline{java.util.concurrent} ~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}. Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeted at an implementation language (C) rather than a toy language or algorithm-level representation. %do we want to say this part?

@@

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
		%		\label{traverse_lock_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\begin{lstlisting}[language=caml, style=myStyle]
			let insert r k =
			  lockNode r;
			  let n = traverse r k in
			  let res = insertOp n k in
			  unlockNode n; 
			  res\end{lstlisting}
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		%		\label{traverse_lock_b}
	\end{subfigure}
	\caption{An example search structure template (lock coupling)}
	\label{template-ex}
\end{figure}

@@

Logical atomicity is a separation logic technique for concisely specifying the behavior of a concurrent operation. A logically atomic triple has the form $\fforall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread. More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$. Its concurrent counterpart could be specified as 
\begin{mathpar}
	\fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \\ \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the logically atomic specifications shown in Figure~\ref{search-specs}. In other words, each operation takes effect atomically on an abstract map $m$, reading and/or updating it as appropriate to the operation.

\begin{figure*}[!ht]
	\centering
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
	\\
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{lookup(r, k)}\ 
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
	\\
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{delete(r, k)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
\caption{The logically atomic specifications for search structure operations}
\label{search-specs}
\end{figure*}

@@

This is the theory behind the template approach to verification. Unfortunately, it is not realized in practice. In the Coq development of Krishna et al., different templates use slightly different specifications for data structure operations and \lstinline{findNext}; the specifications proved for data structure implementations do not always match those used by templates (which is possible because they use separate provers for the sequential and concurrent proofs); and some data structures rely on other functions that do not appear in the templates (e.g., maintenance operations that split full nodes in B-trees). In the following presentation, we follow the template plan as far as possible, noting the places where we must deviate from the uniform approach; we discuss maintenance operations in further detail in Section~\ref{internal_reorganize}.

@@

The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The specification of \texttt{insert} is:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\\ \texttt{insert(r, k, v)}\  \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the update to the concrete data structure corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the data structure.
Finally, we release the lock acquired by \lstinline{traverse} and deallocate the \lstinline{pn} structure.

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
\end{mathpar}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

@@

\subsection{Templates and Internal Reorganization}
\label{internal_reorganize}
The \lstinline{delete} operation on binary search trees highlights a non-obvious aspect of the template approach. The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}). In most of the data structures verified by Krishna et al., deletion can decomposed similarly, by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation. When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
rotating three-node sections of the tree until the node to be deleted is at a leaf. This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them).

Krishna et al. also discuss an operation of this sort, the \lstinline{split} operation in B-link trees. Their approach is to assume the existence of a separate maintenance thread that constantly crawls the tree searching for full nodes and splitting them. The split itself is a no-op on the abstract state of the data structure: it makes internal structural changes but does not change the set of keys stored in the tree. This complicates the description of the template approach, but such a maintenance operation is necessary for many concurrent data structures. In general, we believe that the template approach should be thought of as the combination of \emph{three} kinds of operations: concurrent, data-structure-agnostic template operations; sequential data structure operations; and no-op maintenance operations, which are both concurrency-aware and data-structure-specific, but whose specifications leave the abstract state unchanged. For instance, rotation in the BST can be seen as a maintenance operation, and the deletion of a node that has already been rotated to a leaf position is a sequential, data-structure-specific operation. Rebalancing operations in AVL and red-black trees would also be considered maintenance operations. This can be seen as a limitation of the template approach, since concurrency cannot be completely decoupled from data-structure-specific reasoning, but because maintenance operations are logical no-ops, they should be easier to verify than concurrent data structure operations in general. In future work, we hope to provide a formal account of these three categories of operations, and a theorem describing how the proofs of each kind of operation combine to yield a top-level verified data structure. This will be essential for applying the template approach to the complexity of real-world high-performance data structure implementations.

>>>>

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{Abs}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Ref }}}
\newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
\newcommand{\infp}{\ensuremath{\mathsf{Ref }}}

@@

Our specific contributions are:
\begin{itemize}
	\item We reimplement the template approach without flow interfaces, with a simple interface involving only the concept of ``keys belonging in this node/subtree''.
	\item We implement the template approach in VST, allowing us to apply it to C programs and obtain end-to-end correctness proofs in a single verification system.
	\item To the best of our knowledge, this is the first mechanized verification of a template approach to concurrent data structure implementations in a real-world programming language, and its first application to data structures not written specifically as case studies.
	\item We give a precise description of search structure templates, and identify places where it is difficult in practice to preserve the boundary between template and data structure.
\end{itemize}

\subsection*{Related Work}
Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}. The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach. Some of the more complex data structures verified include Java's \lstinline{ConcurrentSkipListMap}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}. Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeting an implementation language (C) rather than a toy language or algorithmic representation. %do we want to say this part?

@@

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
		%		\label{traverse_lock_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\begin{lstlisting}[language=caml, style=myStyle]
			let insert r k =
			  lockNode r;
			  let n = traverse r k in
			  let res = insertOp n k in
			  unlockNode n; 
			  res\end{lstlisting}
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		%		\label{traverse_lock_b}
	\end{subfigure}
	\caption{The lock-coupling search structure template}
	\label{template-ex}
\end{figure}

@@

Logical atomicity is a separation logic technique for concisely specifying the behavior of a concurrent operation. A logically atomic triple has the form $\fforall a.\,\left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread. More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
\[\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}\] Its concurrent counterpart could be specified as 
\begin{mathpar}
	\fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\  \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{lookup(r, k)}\ \\
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
\end{mathpar}
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{delete(r, k)}\ \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation. Krishna et al. use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.

@@

This is the theory behind the template approach to verification. Unfortunately, it is not straightforward to realize in practice. In the Coq development of Krishna et al., different templates use slightly different specifications for data structure operations and \lstinline{findNext}; the specifications proved for data structure implementations do not always match those used by templates (which is possible because they use separate provers for the sequential and concurrent proofs); and some data structures rely on other functions that do not appear in the templates (e.g., maintenance operations that split full nodes in B-trees). In the following presentation, we follow the template plan as far as possible, noting the places where we must deviate from the uniform approach; we discuss maintenance operations in further detail in Section~\ref{internal_reorganize}.

@@

The \lstinline{insert} function uses this specification of \lstinline{traverse} to update the state of the data structure. The desired specification of \texttt{insert} is:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\  \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
where $m$ and $m[\texttt{k} \mapsto \texttt{v}]$ denote the tree's abstract states before and after the function's execution, respectively. The proof of \lstinline{insert} is outlined in Appendix \ref{sec:apd_proof}, but informally it is quite simple. We begin by initializing the \lstinline{pn} struct with the root node and acquiring its lock, allowing us to satisfy the precondition of \lstinline{traverse}. If \lstinline{traverse} returns 0, we have found a node with key \lstinline{k}, and all we need to do is update that node's value to \lstinline{v} (line 6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty node with \lstinline{k} in its range. We then call the data-structure-specific \texttt{insertOp} function, which inserts the new key-value pair at the empty node. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that in both cases, the update to the concrete data structure corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the data structure.
Finally, we release the lock acquired by \lstinline{traverse} and deallocate the \lstinline{pn} structure.

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
\end{align*}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

@@

\subsection{Templates and Internal Reorganization}
\label{internal_reorganize}
The \lstinline{delete} operation on binary search trees highlights a limitation of the template approach as we have described it. The \lstinline{insert} and \lstinline{lookup} operations can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insertOp} or \lstinline{lookupOp}). In most of the prior template examples, deletion can decomposed similarly, by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation. When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
rotating three-node sections of the tree until the node to be deleted is at a leaf. This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them).

Krishna et al. also discuss an operation of this sort, the \lstinline{split} operation in B-link trees. Their approach is to assume the existence of a separate maintenance thread that constantly crawls the tree searching for full nodes and splitting them; any operation that cannot take effect on full nodes can simply retry until the target node is split. The split itself is a no-op on the abstract state of the data structure: it makes internal structural changes but does not change the set of keys stored in the tree. Maintenance operations of this sort appear in almost every reasonably complex concurrent data structure, and so the template approach must account for their existence. In other words, in general the template approach must divide the operations of a target data structure into three categories: concurrent, data-structure-agnostic template operations, which can be verified once and applied to multiple data structures; sequential data structure operations, which can be verified once and plugged into multiple templates; and no-op maintenance operations, which are both concurrency-aware and data-structure-specific, but whose specifications leave the abstract state unchanged. For instance, rotation in the BST can be seen as a maintenance operation, and the deletion of a node that has already been rotated to a leaf position is a sequential, data-structure-specific operation. Rebalancing operations in AVL and red-black trees would also be considered maintenance operations. The existence of this third category means that we cannot completely separate concurrency reasoning from data structure reasoning, but because maintenance operations are logical no-ops, they should be easier to verify in a non-decomposed style than most concurrent data structure operations.
